---
html_document: default
author: "Joshua Chu"
date: "10/23/2020"
line: null
pdf_document: default
subtitle: Project Supplemental Material
title: 'DSC607: Intro to Data Mining'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

R Markdown was used to author this HTML document using a dataset obtain from
baseball savant (MLB, 2020).

## Project Description

Major League Baseball pitchers can manipulate the movement of a ball by increasing
or decreasing the spin rate of a baseball (Nagami, 2013). There are three forces
that influence ball movement: drag force due to air movement, gravity, and the
Magnus force, which the produces vertical and lateral ball movement (Nagami, 2013).
Since on any given day a pitcher has no control on how the gravity impacts 
or influences ball movement, manipulating the spin rate of a baseball can
directly increase or decrease Magnus force.

Evaluating the spin rates of different pitch types for Major League Baseball 
pitchers may reveal novel relationships between spin rates and resulting metrics
such as total number of strikeouts, earned runs allowed, strikeout percentage,
and more. The following analysis will examine the spin rates of baseballs and
the relationship with strikeout percentage.


## Required Packages

The following packages are required for data manipulation (dplyr), constructing the
required visuals (ggplot2, cowplot, corrplot), and modeling the data using a
Bayesian Network model (e1071, factoextra, tidyverse, caret).

```{r message = FALSE, warning=FALSE}
# load the required libraries for project
library(dplyr); library(ggplot2); library(cowplot); library(corrplot); 
library(ggpubr); library(factoextra); library(e1071); library(tidyverse);
library(caret)
```


## Data Loading and Summary

The dataset was obtained from Baseball Savant, which is a website that provides
standard and advance baseball statistics accumulated since 1950 and 2015,
respectively (Savant, 2020).

```{r}
# remove variables saved in environment
rm(list = ls())

# load dataset and rename columns
x <- read.csv("stats.csv")

# summary statistics for the dataset
summary(x)
```

The data is comprised of 767 observations and 36 variables for Major League 
Baseball pitchers for the year 2019 (Savant, 2020). Below shows the features of
interest for this investigation. There are multiple missing values for the four
pitch types and will need to be resolved.

     1. p_k_percent: strikeout percentage
     2. sl_avg_spin: average slider spin rate (revolutions per minute)
     3. ch_avg_spin: average changeup spin rate (revolutions per minute)
     4. cu_avg_spin: average curveball spin rate (revolutions per minute)
     5. fastball_avg_spin: average fastball spin rate (revolutions per minute)


## Exploratory Data Analysis

A correlation plot was used to examine the relationship between the pitch types
and strikeout percentage, as well as additional features that may reveal novel
relationships.

```{r}
# subset numerical data from categorical
y <- x[,c(12:19,22:35)]

# correlation table
corrplot(cor(y[,c(1:4,11:22)],use = "complete.obs"), type = "upper", 
         tl.col = "black",tl.srt = 45, tl.cex = 1, cl.cex = 1, 
         col=colorRampPalette(c("blue","white","red"))(200))
```

A positive relationship exists between the spin rates and the total number of 
strikeouts and strikeout percentage. While interesting, the percent strikeout 
metric is more valuable because the frequency includes the total number of
batters a pitcher faces throughout the season (Sawchik, 2018). For instance, a
strikeout total of 138 may not seem impressive since many pitchers accumulate 
over 150 strikeouts for an entire season. However, examining the strikeout 
percentage for the same player, which is 47.8% and the best mark in MLB for 2019,
we see this pitcher is highly efficient since his strikeout percentage is a much
higher frequency than the MLB average (22%; summary stats shown above). Therefore,
the strikeout percentage and four pitch types will be utilized for further
investigation. To confirm a linear relationship, a linear regression between the
strikeout percentage and pitch types were performed.

```{r message = FALSE}
# linear regressions to confirm relationship between strikeout percentage
# and the four different pitch types
a <- ggplot(y,aes(sl_avg_spin,p_k_percent))+geom_point(na.rm = TRUE) +
        stat_smooth(method = "lm", se = FALSE, na.rm = TRUE, color = "red", size = 1.5) +
        stat_cor(label.x = 1750, label.y = 47, na.rm = TRUE, size = 3) +
        stat_regline_equation(label.x = 1750, label.y = 44, na.rm = TRUE, size = 3) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

b <- ggplot(y,aes(ch_avg_spin,p_k_percent))+geom_point(na.rm = TRUE) +
        stat_smooth(method = "lm", se = FALSE, na.rm = TRUE, color = "red", size = 1.5) +
        stat_cor(label.x = 2000, label.y = 45, na.rm = TRUE, size = 3) +
        stat_regline_equation(label.x = 2000, label.y = 42, na.rm = TRUE, size = 3) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

c <- ggplot(y,aes(cu_avg_spin,p_k_percent))+geom_point(na.rm = TRUE) +
        stat_smooth(method = "lm", se = FALSE, na.rm = TRUE, color = "red", size = 1.5) +
        stat_cor(label.x = 1000, label.y = 45, na.rm = TRUE, size = 3) +
        stat_regline_equation(label.x = 1000, label.y = 42, na.rm = TRUE, size = 3) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

d <- ggplot(y,aes(fastball_avg_spin,p_k_percent))+geom_point(na.rm = TRUE) +
        stat_smooth(method = "lm", se = FALSE, na.rm = TRUE, color = "red", size = 1.5) +
        stat_cor(label.x = 1750, label.y = 45, na.rm = TRUE, size = 3) +
        stat_regline_equation(label.x = 1750, label.y = 42, na.rm = TRUE, size = 3) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

plot_grid(a,b,c,d)
```

According to the regression analysis, all regression models were statistically
significant with a p value < 0.05. The coefficient of determination varies
among the models, with fastballs performing the best and the changeup performing
the worst among all pitch types. Now that I confirmed a relationship exists, we
can move on to using the dataset to categorize pitchers according to this
relationship.

```{r}
# subset dataset by pitch spin rates
df  <- x[,c(14,26,29,32,35)]

# summary statistics for the dataset
summary(df)
```

The original dataset was further partitioned to only include the strikeout percentage
and spin rates for the aforementioned pitch types. Spin rates are of particular 
interest to pitchers because the 'movement' of a ball can be altered by different
spin rates, which in turn could impact a batters perception for the balls 
trajectory, inducing more swing and misses, and generating a large strikeout
percentage (Nagami, 2013).

Again, examining the summary statistics reveals missing values are still 
prevalent for most features in the dataset with the strikeout percentage variable
being the exception. Changeups, on average, possess the lowest spin rates for 
all pitch types, and the curveball and slider have the fastest. The minimum 
value for the changeup is much smaller than the other pitch types. In all cases,
the median and mean do not differ by much, indicating there are few to no 
outliers present in the dataset.

```{r}
# histogram for strikeout percentage
ggplot(df,aes(p_k_percent)) + geom_histogram(color="black", fill="white", bins = 40,na.rm = TRUE) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))
```

The strikeout percentage attribute has a normal distribution, with an odd
frequency at the zero marker. This likely indicates there are multiple picthers
with a zero strikeout percentage. Because this feature is normally distributed, 
no other manipulations to this variable will be performed. Since the data are
normally distributed, I could use the quartiles identified in the summary stats
to assign pitchers to categories, but this will be investigated later.

```{r}
# examine spin rates for pitches
q <- ggplot(df,aes(sl_avg_spin))+geom_histogram(color="black", fill="white", bins = 40, na.rm = TRUE) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

r <- ggplot(df,aes(ch_avg_spin))+geom_histogram(color="black", fill="white", bins = 40, na.rm = TRUE) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

s <- ggplot(df,aes(cu_avg_spin))+geom_histogram(color="black", fill="white", bins = 40, na.rm = TRUE) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

t <- ggplot(df,aes(fastball_avg_spin))+geom_histogram(color="black", fill="white", bins = 40, na.rm = TRUE) +
        theme_bw()+theme(axis.title = element_text(face = "bold",size = 20),
                         axis.text = element_text(size = 12))

plot_grid(q,r,s,t)
```

Surprisingly, the spin rates for all four pitch types are normally distributed.
Consistent with the summary statistics, the changeup has the smallest mean
compared to the other three pitch types. Visually, the fastball feature appears
to have more data points than the other, however, replacing the NA values with 
the mean of each pitch type should resolve any feature size bias.


## Determine the Number of Pitcher Categories Using Cluster Analysis

```{r}
# eliminate NAs from dataset and replace with 0
df[is.na(df)] <- 0

# convert to dataframe
df  <- data.frame(df)

# scale data
df <- scale(df)
```

As noted earlier, NAs were discovered in the dataset and needed to be removed
prior to the analysis. All missing values were converted to 0 and the dataset 
was formatted to a data frame, followed by scaling the data to make the 
variables comparable.

```{r}
# determine the number of clusters using the elbow method
fviz_nbclust(df, kmeans, method = "wss")
```

The above plot was utilized to identify the optimal number of clusters needed to
perform the K means cluster analysis. df is the data frame the function will use
to determine the number of clusters; kmeans is a partitioning arguement indicating
the cluster method; and the final variable is the method used to estimate the optimal
number of clusters. Using the elbow method, the optimal number of clusters was 
determined to be 5 (Guide, 2020). 

```{r}
# cluster analysis
k <- kmeans(df, centers = 5, nstart = 25)
fviz_cluster(k, data = df)
```

kmeans was the function utilized to construct the cluster plot: df is the data
frame, the number of centers was determine to be 5 (previous step), and the 
nstart indicates the number of random sets that should be attempted and reports
on the 'best' clustering configuration (Guide, 2020).

All clusters but 4, appear to overlap with one or more additional clusters. 
Cluster 2 appears to contain 2 separate clusters within this larger cluster, and
may benefit from sub-clustering. Group 4 is interesting because it appears very 
different from the other groups based on its distance from them, and may 
indicate an outlier group. Further analysis identified this group possessed 
zeros for all pitch types, which indicates the spin rates for all of these 
pitchers did not have any usable data (i.e. all values were NAs). A few of the
records from the original dataset containing the NAs can be observed below.

```{r}
# examine the last 6 rows of the original dataset
tail(x[,c(14,26,29,32,35)])
```

Originally, I thought the dataset had random NAs throughout the features, not
necessarily consecutive records having NAs. Now that these have been identified,
the pitchers that did not contain usable data were removed (records 707 through
767) and the additional missing values will be replaced with the average values 
for their respective features (shown below). The cluster analysis was repeated 
using a center number of four instead of five.

```{r}
# subset dataset by pitch spin rates
df  <- x[1:706,c(14,26,29,32,35)]

# replace NAs with average of each feature
df$sl_avg_spin[is.na(df$sl_avg_spin)] <- 2401
df$ch_avg_spin[is.na(df$ch_avg_spin)] <- 1774
df$cu_avg_spin[is.na(df$cu_avg_spin)] <- 2490
df$fastball_avg_spin[is.na(df$fastball_avg_spin)] <- 2259

# convert to dataframe
df  <- data.frame(df)

# scale data
df1 <- scale(df)
```

```{r}
# cluster analysis
k1 <- kmeans(df1, centers = 4, nstart = 25)
fviz_cluster(k1, data = df1)
```

While the presumed outliers have shifted toward the other data points, there 
remains considerable overlap between all clusters. The poor separation suggests
the Bayesian Network model may produce poor results when attempting to catagorize
pitchers. The next two plots lowered the number of centers to try and reduce the
overlap between clusters.

```{r}
# cluster analysis
k1 <- kmeans(df1, centers = 3, nstart = 25)
fviz_cluster(k1, data = df1)
```

```{r}
# cluster analysis
k1 <- kmeans(df1, centers = 2, nstart = 25)
fviz_cluster(k1, data = df1)
```

Reducing the number of centers progressively improved the clustering of the data
points. While the majority of the data points are located near the center of the
plot, there is a better separation between the clusters than previously seen. To
try and completely separate the clusters, the cmeans function was used to resolve
the 'fuzzy' nature of the current cluster results (Tan, 2018).

```{r}
# use cmeans to resolve the overlap between clusters 1 and 2
k2 <- cmeans(df1, centers = 2)
fviz_cluster(list(data = df1, cluster=k2$cluster), 
             ellipse.type = "norm",
             ellipse.level = 0.85,
             palette = "jco",
             ggtheme = theme_minimal())
```

Using the same number of clusters as before (i.e. k = 2), the cmeans function was
used as an alternative function to resolve the overlap between clusters 1 and 2. 
Additional parameters were used in the cmeans function, however, these parameters
failed to improve the results and were eliminated for simplicity.

The function fviz_cluster was used to visualize the plot using a list object as
the cluster components. The data to be plotted is found in df1, and the clusters
each data point is assigned to is given by the cluster arguement. The next two 
arguements places ellipses around the data points that belong to each cluster,
while enclosing only 85% of the data points. Finally, the last two arguements
are for visualization purposes only. 

Visually comparing this plot with the previous plot (where k = 2), there does
not appear to be any differences or improvements in separating the two clusters.
Overall, the results suggest I should use 2 pitching catagories (e.g. below 
average and above average) to assign pitchers. Increasing the number of 
assignments will likely reduce the accuracy of the model. 


## Prepare Data for Bayesian Network Modeling

Missing values have been addressed and the number of pitching assignments have
been determined. Now that I know the required structure of the dataset, I need
to prepare the data for analysis. The first step is to construct a single pitch
column, which would combine all of pitch types into a single feature.

```{r}
# convert to dataframe
df2  <- data.frame(df)

# prepare dataset for analysis
c  <- df2[,1:2]
c1 <- df2[,c(1,3)]
c2 <- df2[,c(1,4)]
c3 <- df2[,c(1,5)]

c<-mutate(c,"spin_slider")
colnames(c) <- c("percent_strikeouts", "spin_rate", "pitch")
c1<-mutate(c1,"spin_changeup")
colnames(c1) <- c("percent_strikeouts", "spin_rate", "pitch")
c2<-mutate(c2,"spin_curve")
colnames(c2) <- c("percent_strikeouts", "spin_rate", "pitch")
c3<-mutate(c3,"spin_fastball")
colnames(c3) <- c("percent_strikeouts", "spin_rate", "pitch")
e<-rbind(c,c1,c2,c3)
```

The next step is to construct a new column to assign pitching effectiveness. 
This new feature will be called pitch_assignment and will be given a below average
or above average designation based on being below or above the strikeout 
percentage for the particular record. Following this assignment, the strikeout
percentage was removed from the data frame prior to analysis, and all variables
were converted to factors, a necessary transformation when constructing a Bayesian
Network model in R.

```{r}
# assign each pitch to a category; categories were determined using the percent
# strikeout quartiles
e <- mutate(e,pitch_assignment = 0)
e$pitch_assignment[e$percent_strikeouts <= mean(e$percent_strikeouts)] <- "Below_avg"
e$pitch_assignment[e$percent_strikeouts > mean(e$percent_strikeouts)] <- "Above_avg"

# construct final dataset
sub_e <- e[,c(2:4)]
final_e <- sub_e
final_e$spin_rate <- as.factor(final_e$spin_rate)
final_e$pitch <- as.factor(final_e$pitch)
final_e$pitch_assignment <- as.factor(final_e$pitch_assignment)
```

A training and testing set was constructed for the model. Using the createDataPartition
function splits the data into a 85% training and 15% testing set. Once the two
sets were made, basic statistics were recorded for both sets of data.

```{r}
# the createDataPartition does a stratified random split dividing the data into 
# 85% training and 15% testing
trn <- createDataPartition(y = final_e$pitch_assignment, p = 0.85, list = FALSE)
training <- final_e[trn,]
test <- final_e[-trn,]

# determine the number of records for each dataset
c(Training=nrow(training),Test=nrow(test))

# find the number of records for each pitching assignment
c(Training=table(training$pitch_assignment),Test=table(test$pitch_assignment))

# find the percentage of records for each dataset
c(Training=prop.table(table(training$pitch_assignment)),Test=prop.table(table(test$pitch_assignment)))
```

The training and test set contains 2401 and 423 records, respectively, consistent
with an 85% and 15% split. By the numbers, the below average category is more
abundant for the training and testing dataset. Examining the distributions between
the two datasets for each category, the percentages do not vary, indicating there
is not an extreme bias towards one category between the two datasets. This is
important because if the training set would flip the frequencies for each category,
model predictions may be inaccurate due to inconsistencies between the datasets.


## Perform Bayesian Network Modeling

```{r}
# train the model
Naive_Bayes_Model <- naiveBayes(pitch_assignment ~., data=training)


# predicitons
NB_Predictions <- predict(Naive_Bayes_Model,test)
table(NB_Predictions,test$pitch_assignment)
```

According to the table above, predicting an Above_avg pitcher is only 47% accurate
and a Below_avg pitcher increases to 62%. This is considerably better than other
predicted models that I have not shown, however, improvements to the dataset or how
the analysis is performed is required. One immediate idea is gathering strikeout
percentage results for each individual pitch type, rather than using the overall
percentage recorded for a pitcher. Logically this makes sense because each pitcher
favors a specific pitch to strikeout a batter. Including this data into the model
would likely improve the predictive power and accuracy.


## References

Advanced Media, MLB. (2020). Retrieved from https://baseballsavant.mlb.com/leaderboard/custom?year=2019&type=pitcher&filter=&sort=4&sortDir=asc&min=q&selections=xba,xslg,xwoba,xobp,xiso,exit_velocity_avg,launch_angle_avg,barrel_batted_rate,&chart=false&x=xba&y=xba&r=no&chartType=beeswarm. Date Accessed: Sept 27, 2020.

UC Business Analytics R Programming Guide. (2020). K-means Cluster Analysis. Retrieved from https://uc-r.github.io/kmeans_clustering. Accessed Date: Oct 15, 2020.

Nagami, T., Higuchi, T., & Kanosue, K. (2013). How baseball spin influences the performance of a pitcher. The Journal of Physical Fitness and Sports Medicine, 2(1), 63-68. doi:10.7600/jpfsm.2.63

Savant. (2020). Statcast Leaderboards. Retrieved from https://baseballsavant.mlb.com/leaderboard/custom?year=2019&type=pitcher&filter=&sort=4&sortDir=asc&min=q&selections=xba,xslg,xwoba,xobp,xiso,exit_velocity_avg,launch_angle_avg,barrel_batted_rate,&chart=false&x=xba&y=xba&r=no&chartType=beeswarm. Accessed Date: September 20, 2020.

Sawchik, T. (2018). Baseball’s Top Staffs Have Come Around On The High-Spin Fastball. Retrieved from https://fivethirtyeight.com/features/baseballs-top-staffs-have-come-around-on-the-high-spin-fastball/. Accessed Date: Sept 28, 2020.

Tan, P.-N., Steinbach, M., Karpatne, A., & Kumar, V. (2018). Introduction to Data Mining: Pearson; 2nd Edition.